---
title: "Quick Start: VariantAnnotation Function Categories"
author: "Bioconductor Package Maintainers"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
vignette: >
  %\VignetteIndexEntry{Quick Start: VariantAnnotation Function Categories}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: yes
    toc: yes
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Overview

This vignette walks through the main function categories of the
`VariantAnnotation` package using the bundled `chr22.vcf.gz` example file.
For deeper coverage see `vignette("VariantAnnotation")`,
`vignette("filterVcf")`, and `vignette("ensemblVEP")`.

```{r load}
library(VariantAnnotation)
library(IRanges)
fl <- system.file("extdata", "chr22.vcf.gz", package = "VariantAnnotation")
```

# VCF I/O

Read, write, scan, and filter VCF files. `ScanVcfParam` controls which
chromosomal regions, samples, and fields are loaded.

```{r io}
vcf <- readVcf(fl, genome = "hg19")
vcf

dest <- tempfile(fileext = ".vcf")
writeVcf(vcf, dest)

## Filter to variants with non-missing QUAL > 20
filt_dest <- tempfile(fileext = ".vcf.gz")
filterVcf(fl, genome = "hg19", destination = filt_dest,
          filters = FilterRules(list(
              minQ = function(x) !is.na(fixed(x)$QUAL) & fixed(x)$QUAL > 20)))
```

Key functions: `readVcf`, `writeVcf`, `scanVcf`, `scanVcfHeader`,
`filterVcf`, `indexVcf`, `VcfFile`, `ScanVcfParam`

# VCF Data Structures and Accessors

`CollapsedVCF` (one row per variant, multi-allelic ALT stored as list) and
`ExpandedVCF` (one row per allele) form the core data model.

```{r accessors}
ref(vcf)[1:4]
alt(vcf)[1:4]
info(vcf)[1:4, 1:4]
geno(vcf)$GT[1:4, 1:3]
header(vcf)
evcf <- VariantAnnotation::expand(vcf)
evcf
```

Key functions/classes: `VCF`, `CollapsedVCF`, `ExpandedVCF`, `VCFHeader`,
`fixed`, `ref`, `alt`, `qual`, `filt`, `info`, `geno`, `header`, `meta`,
`vcfFields`, `expand`

# Variant Classification

Classify variants by mutation type. All functions operate on `VCF` or
`VRanges` objects and return logical vectors.

```{r classify}
table(isSNV(vcf))
table(isIndel(vcf))
table(isTransition(vcf))
table(isSubstitution(vcf))
```

Key functions: `isSNV`, `isInsertion`, `isDeletion`, `isIndel`, `isDelins`,
`isTransition`, `isSubstitution`; `VariantType` class hierarchy
(`CodingVariants`, `SpliceSiteVariants`, `IntronVariants`, `FiveUTRVariants`,
`ThreeUTRVariants`, `IntergenicVariants`, `PromoterVariants`)

# Functional Annotation

Map variants to transcripts and predict protein-level consequences.
The example VCF uses NCBI-style seqnames ("22") while the TxDb uses UCSC
style ("chr22"), so we harmonise with `seqlevelsStyle`.

```{r functional}
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
library(BSgenome.Hsapiens.UCSC.hg19)
library(GenomeInfoDb)

txdb <- TxDb.Hsapiens.UCSC.hg19.knownGene
seqlevelsStyle(vcf) <- "UCSC"   # "22" -> "chr22" to match TxDb

loc <- locateVariants(vcf, txdb, region = CodingVariants())
head(loc, 4)

coding <- predictCoding(vcf, txdb, seqSource = Hsapiens)
head(coding, 4)
```

Key functions: `locateVariants`, `predictCoding`, `summarizeVariants`,
`getTranscriptSeqs`

# Genotype Analysis

Convert VCF genotype calls to a `SnpMatrix` for downstream association
testing with the **snpStats** package.

```{r geno}
library(snpStats)
seqlevelsStyle(vcf) <- "NCBI"   # restore "22" for consistency
sm <- genotypeToSnpMatrix(vcf)
head(col.summary(sm$genotypes), 4)
```

Key functions: `genotypeToSnpMatrix`, `GLtoGP`, `probabilityToSnpMatrix`

# Variant Ranges (VRanges)

`VRanges` extends `GRanges` with per-variant fields: alleles, read depth,
and soft/hard filter results.

```{r vranges}
vr <- VRanges(seqnames = "chr1", ranges = IRanges(1000, 1000),
              ref = "A", alt = "T",
              totalDepth = 30, altDepth = 15,
              sampleNames = "sample1")
altFraction(vr)
asVCF(vr)
```

Key functions/classes: `VRanges`, `VRangesList`, `makeVRangesFromGRanges`,
`totalDepth`, `altDepth`, `refDepth`, `altFraction`, `softFilterMatrix`,
`hardFilters`, `called`, `asVCF`, `stackSamples`

# Ensembl VEP REST API

Query the Ensembl Variant Effect Predictor via its REST endpoint for
human variants (hg38 / GRCh38). A live internet connection is required;
the chunk is skipped gracefully if the endpoint is unreachable.

```{r vep}
vep_ok <- tryCatch({
    resp <- post_Hs_region(chr = "7", pos = 155800001,
                           id = "myVar", ref = "A", alt = "T")
    httr::status_code(resp) == 200L
}, error = function(e) FALSE)

if (vep_ok) {
    vcf_vep <- readVcf(fl)
    hits <- vep_by_region(vcf_vep[1:10], snv_only = FALSE, chk_max = FALSE)
    cat("VEP response status:", httr::status_code(hits), "\n")
    result <- jsonlite::fromJSON(jsonlite::toJSON(httr::content(hits)))
    print(head(result[, c("id", "most_severe_consequence")], 4))
} else {
    message("Ensembl VEP REST endpoint not reachable; skipping.")
}
```

Key functions: `post_Hs_region`, `vep_by_region`

# Session Information

```{r sessionInfo}
sessionInfo()
```
